---
title: "A Replication of Karlan and List (2007)"
author: "Minh Vo"
date: 2025-04-20
callout-appearance: minimal # this hides the blue "i" icon on .callout-notes
format: html
---


## Introduction

Dean Karlan at Yale and John List at the University of Chicago conducted a field experiment to test the effectiveness of different fundraising letters. They sent out 50,000 fundraising letters to potential donors, randomly assigning each letter to one of three treatments: a standard letter, a matching grant letter, or a challenge grant letter. They published the results of this experiment in the _American Economic Review_ in 2007. The article and supporting data are available from the [AEA website](https://www.aeaweb.org/articles?id=10.1257/aer.97.5.1774) and from Innovations for Poverty Action as part of [Harvard's Dataverse](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/27853&version=4.2).

In this experiment, Dean Karlan and John List wanted to understand what makes people more likely to donate to a cause, specifically, whether offering to match donations actually works. They mailed out 50,000 fundraising letters to past donors of a nonprofit, randomly assigning each recipient to one of a few different versions. Some people got a standard letter, while others were told their donation would be matched — either dollar-for-dollar (1:1), doubled (2:1), or tripled (3:1). A few versions also included different thresholds for how much the matching donor would contribute in total. By comparing how people responded across these groups, the researchers could see not just whether matching worked, but whether the size or structure of the match made a difference. As it would turn out, even just offering a match, any match, made people more likely to give.

This project seeks to replicate their results.


## Data

### Description

We begin by loading the dataset using the haven package, which allows us to read Stata .dta files into R. The dataset contains 50,083 observations, each representing an individual who received a fundraising letter. It includes information on treatment assignment, match ratios, suggested donation amounts, previous giving behavior, and demographic characteristics.

```{r}
library(haven)
library(dplyr)
library(ggplot2)

# Load data
df <- read_dta("../../data/karlan_list_2007.dta")

df %>%
  select(treatment, gave, amount, mrm2, female, ask, hpa) %>%
  summary()

```

### Variable Definitions

| Variable             | Description                                                         |
|----------------------|---------------------------------------------------------------------|
| `treatment`          | Treatment                                                           |
| `control`            | Control                                                             |
| `ratio`              | Match ratio                                                         |
| `ratio2`             | 2:1 match ratio                                                     |
| `ratio3`             | 3:1 match ratio                                                     |
| `size`               | Match threshold                                                     |
| `size25`             | \$25,000 match threshold                                            |
| `size50`             | \$50,000 match threshold                                            |
| `size100`            | \$100,000 match threshold                                           |
| `sizeno`             | Unstated match threshold                                            |
| `ask`                | Suggested donation amount                                           |
| `askd1`              | Suggested donation was highest previous contribution                |
| `askd2`              | Suggested donation was 1.25 x highest previous contribution         |
| `askd3`              | Suggested donation was 1.50 x highest previous contribution         |
| `ask1`               | Highest previous contribution (for suggestion)                      |
| `ask2`               | 1.25 x highest previous contribution (for suggestion)               |
| `ask3`               | 1.50 x highest previous contribution (for suggestion)               |
| `amount`             | Dollars given                                                       |
| `gave`               | Gave anything                                                       |
| `amountchange`       | Change in amount given                                              |
| `hpa`                | Highest previous contribution                                       |
| `ltmedmra`           | Small prior donor: last gift was less than median \$35              |
| `freq`               | Number of prior donations                                           |
| `years`              | Number of years since initial donation                              |
| `year5`              | At least 5 years since initial donation                             |
| `mrm2`               | Number of months since last donation                                |
| `dormant`            | Already donated in 2005                                             |
| `female`             | Female                                                              |
| `couple`             | Couple                                                              |
| `state50one`         | State tag: 1 for one observation of each of 50 states; 0 otherwise  |
| `nonlit`             | Nonlitigation                                                       |
| `cases`              | Court cases from state in 2004-5 in which organization was involved |
| `statecnt`           | Percent of sample from state                                        |
| `stateresponse`      | Proportion of sample from the state who gave                        |
| `stateresponset`     | Proportion of treated sample from the state who gave                |
| `stateresponsec`     | Proportion of control sample from the state who gave                |
| `stateresponsetminc` | stateresponset - stateresponsec                                     |
| `perbush`            | State vote share for Bush                                           |
| `close25`            | State vote share for Bush between 47.5% and 52.5%                   |
| `red0`               | Red state                                                           |
| `blue0`              | Blue state                                                          |
| `redcty`             | Red county                                                          |
| `bluecty`            | Blue county                                                         |
| `pwhite`             | Proportion white within zip code                                    |
| `pblack`             | Proportion black within zip code                                    |
| `page18_39`          | Proportion age 18-39 within zip code                                |
| `ave_hh_sz`          | Average household size within zip code                              |
| `median_hhincome`    | Median household income within zip code                             |
| `powner`             | Proportion house owner within zip code                              |
| `psch_atlstba`       | Proportion who finished college within zip code                     |
| `pop_propurban`      | Proportion of population urban within zip code                      |


### Balance Test 

As an ad hoc test of the randomization mechanism, I provide a series of tests that compare aspects of the treatment and control groups to assess whether they are statistically significantly different from one another.

To assess the success of the randomization, we compare the treatment and control groups on a pre-treatment variable: mrm2, which represents the number of months since the last donation. If the randomization was effective, we would expect no significant differences between the two groups on this variable.

# T-test for balance
```{r}
t.test(mrm2 ~ treatment, data = df)
```

The two-sample t-test yields a p-value of 0.9049, suggesting that there is no statistically significant difference between the treatment and control groups in terms of their recency of last donation.

# Regression for balance
```{r}
summary(lm(mrm2 ~ treatment, data = df))
```

The linear regression confirms this result. The estimated coefficient on treatment is very small (≈ 0.014) and not statistically significant (p = 0.905), indicating that group assignment does not explain any meaningful variation in donation recency.

These results align with the balance checks shown in Table 1 of the original paper, which are included to assure readers that the random assignment created comparable groups. This gives us confidence that any differences observed in donation behavior later on can be attributed to the treatment itself.

## Experimental Results

### Charitable Contribution Made

First, I analyze whether matched donations lead to an increased response rate of making a donation. 

```{r}
# Proportion barplot
prop_data <- df %>%
  group_by(treatment) %>%
  summarise(rate = mean(gave))

ggplot(prop_data, aes(x = factor(treatment), y = rate, fill = factor(treatment))) +
  geom_bar(stat = "identity") +
  labs(x = "Group (0 = Control, 1 = Treatment)", y = "Proportion Donated", fill = "Group") +
  ggtitle("Donation Rate by Treatment Group") +
  scale_fill_manual(values = c("#FF6F61", "#00BFC4")) +
  theme_minimal()
```
This visual shows that the donation rate was noticeably higher among those who received a matching offer.

### T-test and Linear Regression

We now formally test the difference in donation rates using both a two-sample t-test and a simple linear regression.

```{r}
# T-test
t.test(gave ~ treatment, data = df)

# Linear regression
summary(lm(gave ~ treatment, data = df))
```

Both approaches yield consistent results. The t-test produces a statistically significant p-value (~0.0013), and the regression confirms that being in the treatment group increases the likelihood of donating by about 0.4 percentage points. While this may seem small, it’s meaningful given the very low baseline donation rate (~1.8% in the control group). These results replicate Table 2a Panel A in the original paper and demonstrate that even a modest match offer has a real behavioral effect.

### Probit Regression
To further validate the result, we run a probit regression where the dependent variable is gave and the independent variable is treatment.

```{r}
summary(glm(gave ~ treatment, family = binomial(link = "probit"), data = df))
```

The probit model returns a positive and statistically significant coefficient on the treatment variable, indicating that the matching offer significantly increases the probability of donating. This aligns with Table 3, Column 1 of the original study.

### Differences between Match Rates

In this section, we explore whether the size of the matching ratio affects the likelihood of giving. Specifically, we test whether individuals were more likely to donate when offered a 2:1 or 3:1 match compared to a standard 1:1 match.

### T-tests Between Match Ratios

We begin by running two pairwise t-tests:
- The first compares donation rates between those offered a 1:1 match and those offered a 2:1 match.
- The second compares 2:1 and 3:1 match groups.

```{r}
# 1:1 vs 2:1
t.test(df$gave[df$ratio2 == 1], df$gave[df$ratio == 1])

# 2:1 vs 3:1
t.test(df$gave[df$ratio3 == 1], df$gave[df$ratio2 == 1])
```

The p-values for both comparisons (0.33 and 0.96, respectively) suggest that differences in match ratios do not produce statistically significant changes in donation behavior. This supports the authors’ observation in the paper that while match presence boosts donations, increasing the match size offers diminishing returns.

### Regression on Match Ratios

We also run a regression using indicator variables for 2:1 and 3:1 matches (with 1:1 match as the omitted baseline):

```{r}
summary(lm(gave ~ ratio2 + ratio3, data = df))
```

The regression shows that being offered a 2:1 match increases the likelihood of giving by ~0.36 percentage points compared to a 1:1 match, while a 3:1 match increases it by ~0.37 points. Both coefficients are statistically significant at the 5% level. However, the practical effect sizes are small, and the increase from 2:1 to 3:1 is negligible — suggesting diminishing marginal returns to match size.

```{r}
# Mean donation rates
mean_1to1 <- mean(df$gave[df$ratio == 1])
mean_2to1 <- mean(df$gave[df$ratio2 == 1])
mean_3to1 <- mean(df$gave[df$ratio3 == 1])

# Differences
diff_2vs1 <- mean_2to1 - mean_1to1
diff_3vs2 <- mean_3to1 - mean_2to1
```

These calculations confirm the regression results. The response rate increases slightly from 1:1 to 2:1, and barely at all from 2:1 to 3:1. This aligns with the authors' interpretation on page 8, where they suggest that the presence of a match matters more than its size.

### Interpretation
The evidence here suggests that once donors are presented with a match, increasing the match ratio (from 1:1 to 2:1 to 3:1) does not substantially change behavior. This has important implications for nonprofits: just offering a match may be sufficient, and investing in larger match ratios might not yield proportionally better outcomes.

### Difference in Response Rates
Finally, we compare raw response rates directly:

### Size of Charitable Contribution

In this subsection, I analyze the effect of the size of matched donation on the size of the charitable contribution.

_todo: Calculate a t-test or run a bivariate linear regression of the donation amount on the treatment status. What do we learn from doing this analysis?_

_todo: now limit the data to just people who made a donation and repeat the previous analysis. This regression allows you to analyze how much respondents donate conditional on donating some positive amount. Interpret the regression coefficients -- what did we learn? Does the treatment coefficient have a causal interpretation?_ 

_todo: Make two plot: one for the treatment group and one for the control. Each plot should be a histogram of the donation amounts only among people who donated. Add a red vertical bar or some other annotation to indicate the sample average for each plot._

```{r}
# Bivariate regression
summary(lm(amount ~ treatment, data = df))

# Filter for donors only
df_givers <- df %>% filter(gave == 1)

# Regression on amount among givers
summary(lm(amount ~ treatment, data = df_givers))

# Histogram for treatment
ggplot(df_givers %>% filter(treatment == 1), aes(x = amount)) +
  geom_histogram(binwidth = 10, fill = "steelblue") +
  geom_vline(aes(xintercept = mean(amount)), color = "red", linetype = "dashed") +
  labs(title = "Donation Amounts (Treatment Group)", x = "Amount", y = "Count") +
  theme_minimal()

# Histogram for control
ggplot(df_givers %>% filter(treatment == 0), aes(x = amount)) +
  geom_histogram(binwidth = 10, fill = "darkorange") +
  geom_vline(aes(xintercept = mean(amount)), color = "red", linetype = "dashed") +
  labs(title = "Donation Amounts (Control Group)", x = "Amount", y = "Count") +
  theme_minimal()
```

## Simulation Experiment

As a reminder of how the t-statistic "works," in this section I use simulation to demonstrate the Law of Large Numbers and the Central Limit Theorem.

Suppose the true distribution of respondents who do not get a charitable donation match is Bernoulli with probability p=0.018 that a donation is made. 

Further suppose that the true distribution of respondents who do get a charitable donation match of any size  is Bernoulli with probability p=0.022 that a donation is made.

### Law of Large Numbers

_to do:  Make a plot like those on slide 43 from our first class and explain the plot to the reader. To do this, you will simulate 100,00 draws from the control distribution and 10,000 draws from the treatment distribution. You'll then calculate a vector of 10,000 differences, and then you'll plot the cumulative average of that vector of differences. Comment on whether the cumulative average approaches the true difference in means._

```{r}
set.seed(123)
ctrl <- rbinom(100000, 1, 0.018)
treat <- rbinom(10000, 1, 0.022)
diffs <- treat - sample(ctrl, 10000)
cum_avg <- cumsum(diffs)/seq_along(diffs)

plot(cum_avg, type = "l", col = "blue", ylab = "Cumulative Avg Diff", main = "LLN: Cumulative Difference", xlab = "Simulations")
abline(h = 0.004, col = "red", lty = 2)
```

### Central Limit Theorem

_to do: Make 4 histograms like those on slide 44 from our first class at sample sizes 50, 200, 500, and 1000 and explain these plots to the reader. To do this for a sample size of e.g. 50, take 50 draws from each of the control and treatment distributions, and calculate the average difference between those draws. Then repeat that process 999 more times so that you have 1000 averages. Plot the histogram of those averages. Comment on whether zero is in the "middle" of the distribution or whether it's in the "tail."_

```{r}
sample_sizes <- c(50, 200, 500, 1000)

par(mfrow = c(2, 2))
for (n in sample_sizes) {
  diffs <- replicate(1000, {
    m_treat <- mean(rbinom(n, 1, 0.022))
    m_ctrl <- mean(rbinom(n, 1, 0.018))
    m_treat - m_ctrl
  })
  hist(diffs, main = paste("Sample size =", n), xlab = "Mean Differences")
}
par(mfrow = c(1, 1))
```
