---
title: "Machine Learning"
author: "Minh Vo"
date: 2025-06-11
---

```{r, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)

# Load the dataset
drivers <- read_csv("../../data/data_for_drivers_analysis.csv")
penguins <- read_csv("../../data/palmer_penguins.csv")
yogurt <- read_csv("../../data/yogurt_data.csv")
```

## 1a. K-Means

We implement the K-means clustering algorithm from scratch to cluster penguins based on two continuous features: bill length and flipper length. K-means attempts to partition data into `k` clusters by iteratively updating cluster centers and reassigning points based on Euclidean distance.
```{r, message=FALSE, warning=FALSE}
# Prep the penguins dataset
penguins_clean <- penguins %>%
  select(bill_length_mm, flipper_length_mm) %>%
  na.omit()

set.seed(1)

# Initialize K-means from scratch
k_means_custom <- function(data, k = 3, max_iter = 10) {
  n <- nrow(data)
  centers <- data[sample(1:n, k), ]  # random initial centers
  clusters <- rep(0, n)
  history <- list()

  for (iter in 1:max_iter) {
    # Assign points to closest center
    dists <- sapply(1:k, function(i) rowSums((data - centers[i, ])^2))
    new_clusters <- apply(dists, 1, which.min)

    # Store for plotting
    history[[iter]] <- list(centers = centers, clusters = new_clusters)

    # Update centers
    for (i in 1:k) {
      if (sum(new_clusters == i) > 0) {
        centers[i, ] <- colMeans(data[new_clusters == i, , drop = FALSE])
      }
    }

    # Check convergence
    if (all(new_clusters == clusters)) break
    clusters <- new_clusters
  }

  return(list(final_clusters = clusters, centers = centers, history = history))
}

# Run the algorithm
X <- as.matrix(penguins_clean)
k_result <- k_means_custom(X, k = 3, max_iter = 10)
```

### Optimal Number of Clusters

To determine the appropriate number of clusters for the penguins data, we compute two metrics:
- **Within-cluster sum of squares (WCSS)**: Measures compactness of clusters.
- **Silhouette score**: Measures separation between clusters.

We vary K from 2 to 7 and evaluate both metrics.

```{r, message=FALSE, warning=FALSE}
library(cluster)  # for silhouette()
library(factoextra)  # for easy visualization

# Prepare clean data
X <- na.omit(penguins %>%
               select(bill_length_mm, flipper_length_mm)) %>%
     as.matrix()

# Initialize results
wcss <- c()
silhouette_avg <- c()

# Loop over values of K
for (k in 2:7) {
  k_result <- kmeans(X, centers = k, nstart = 10)
  wcss[k] <- k_result$tot.withinss
  
  # Calculate silhouette score
  sil <- silhouette(k_result$cluster, dist(X))
  silhouette_avg[k] <- mean(sil[, 3])
}

# Plot WCSS
plot(2:7, wcss[2:7], type = "b", col = "steelblue", pch = 19,
     xlab = "Number of Clusters (K)", ylab = "Within-Cluster Sum of Squares",
     main = "Elbow Method: WCSS vs K")

# Plot Silhouette scores
plot(2:7, silhouette_avg[2:7], type = "b", col = "darkorange", pch = 19,
     xlab = "Number of Clusters (K)", ylab = "Average Silhouette Score",
     main = "Silhouette Score vs K")
```




## 2a. K Nearest Neighbors

_todo: use the following code (or the python equivalent) to generate a synthetic dataset for the k-nearest neighbors algorithm.  The code generates a dataset with two features, `x1` and `x2`, and a binary outcome variable `y` that is determined by whether `x2` is above or below a wiggly boundary defined by a sin function._

```{r}
# gen data -----
set.seed(42)
n <- 100
x1 <- runif(n, -3, 3)
x2 <- runif(n, -3, 3)
x <- cbind(x1, x2)

# define a wiggly boundary
boundary <- sin(4*x1) + x1
y <- ifelse(x2 > boundary, 1, 0) |> as.factor()
dat <- data.frame(x1 = x1, x2 = x2, y = y)
```

_todo: plot the data where the horizontal axis is `x1`, the vertical axis is `x2`, and the points are colored by the value of `y`.  You may optionally draw the wiggly boundary._

_todo: generate a test dataset with 100 points, using the same code as above but with a different seed._

_todo: implement KNN by hand.  Check you work with a built-in function -- eg, `class::knn()` or `caret::train(method="knn")` in R, or scikit-learn's `KNeighborsClassifier` in Python._

_todo: run your function for k=1,...,k=30, each time noting the percentage of correctly-classified points from the test dataset. Plot the results, where the horizontal axis is 1-30 and the vertical axis is the percentage of correctly-classified points. What is the optimal value of k as suggested by your plot?_ 



## 2b. Key Drivers Analysis

_todo: replicate the table on slide 75 of the session 5 slides. Specifically, using the dataset provided in the file data_for_drivers_analysis.csv, calculate: pearson correlations, standardized regression coefficients, "usefulness", Shapley values for a linear regression, Johnson's relative weights, and the mean decrease in the gini coefficient from a random forest. You may use packages built into R or Python; you do not need to perform these calculations "by hand."_

_If you want a challenge, add additional measures to the table such as the importance scores from XGBoost, from a Neural Network, or from any additional method that measures the importance of variables._






